The Rockerverse: Images, Packages, and Applications for Containerization with R

### 2020-01-17 manuscript.Rmd
Add a short section on the use of Rocker containers in Dodona

title: "Containers and R---the Rockerverse and beyond"

@@

@@

@@

The [**Jupyter Docker Stacks**](https://github.com/jupyter/docker-stacks/) project are a set of ready-to-run Docker images containing Jupyter applications and interactive computing tools \citep{project_jupyter_jupyter_2018}.
The `jupyter/r-notebook` image includes R and "popular packages", and naturally also the IRKernel ([https://irkernel.github.io/](https://irkernel.github.io/)), an R kernel for Jupyter so that Jupyter Notebooks can contain R code cells. 
R is also included in the catchall `jupyter/datascience-notebook` image\footnote{\href{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}}.
For example, these images allow users to quickly start a Jupyter Notebook server locally or build their own specialised images on top of stable toolsets.
The R installation is installed using the Conda package manager, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See `jupyter/datascience-notebook` `Dockerfile` at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile#L47}.}.

@@

**Gigantum** ([http://gigantum.com/](http://gigantum.com/) is a platform for open and decentralized data science with a focus on using automation and user-friendly tools for easy sharing of reproducible computational workflows.
Gigantum builds on the _Gigantum Client_ (running either locally or on a remote server) for development and execution of data-focused _Projects_, which can be stored and shared via the _Gigantum Hub_ or via a zipfile export.
The Client is a user friendly interface to a backend using Docker containers to package, build, and run Gigantum projects. 
It is configured to use a default set of Docker base images ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)), and users are able to define and configure their own custom images.
The available images include two with R based on Ubuntu Linux have the [`c2d4u` CRAN PPA](https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/) pre-configured for installation of binary R packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
The R images vary in the included authoring environment, i.e., Jupyter in `r-tidyverse` or both Jupyter \& RStudio in `rstudio-server`.
The independent image stack can be traced back to the Gigantum environment and its features.
The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.
The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the `Dockerfile` of `python3-minimal`.} so that appropriate GPU drivers can be enabled automatically when supported.
Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resolution.
For this reason, project repositories store authoritative package information in a separate file per package, allowing Git to directly detect conflicts and changes.
A Dockerfile is generated from this description that inherits from the specified base image, and additional custom Docker instructions may be appended by users, though Gigantum's default base images do not currently include the `littler` tool, which is used by Rocker to install packages within `Dockerfile`s.
Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusion}).
Both approaches enable `apt` as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on plain `Dockerfile`s.
A minor reason in the inception of the images were also project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.

@@

**\pkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) 
<!-- from the CRAN description:
Work with containers over the Docker API. Rather than using system calls to interact with a docker client, using the API directly means that we can receive richer information from docker. The interface in the package is automatically generated using the 'OpenAPI' (a.k.a., 'swagger') specification, and all return values are checked in order to make them type stable.
-->
currently the most powerful Docker client in R.

**\pkg{AzureContainers}** ([https://cran.r-project.org/package=AzureContainers](https://cran.r-project.org/package=AzureContainers)) is an interface to a number of container-related services in Microsoft's [Azure Cloud](https://azure.microsoft.com/), namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/), [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/), and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/).
While it is mainly intended for working with Azure, as a convenience feature it includes lightweight, cross-platform shells to Docker and Kubernetes (tools `kubectl` and `helm`).
These can be used to create and manage arbitrary Docker images and containers, as well as Kubernetes clusters on any platform or cloud service.

@@

R has been historically viewed as a tool for analysis and scientific research--not for creating software that corporations can rely on to continously run in real time.
However, thanks to advancements in R running as a web service, along with with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.

The backbone of this work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package.
The plumber package allows R to be used to host web services through RESTful APIs.
When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.
For instance, if a data scientist at a company creates a machine learning model to tell which customers are likely to make a repeat purchase, if they serve the model through a plumber API then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
Plumber syntax is simple to read and understand: the API is created by commenting above function names in an R script.

@@

@@

- `RSelenium`
- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `gce_vm_template()`)
  - To enable quick deployments of key R services such as RStudio and Shiny onto cloud virtual machines (VMs), this package utilises Dockerfiles to move the labour of setting up those services from the user to a premade Docker image.  For example, by specifying the template `template="rstudio"` in `gce_vm_template()/gce_vm()` an up to date RStudio Service image is launched.  Specifying `template="rstudio-gpu"` will launch an RStudio Server image with a GPU attached, etc.  
- [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
- [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]
- Kubernetes - A popular platform for managing Docker containers is [Kubernetes](https://kubernetes.io/), which is used for a wide variety of containerized applications.  It may be your organisation has a Kubernetes cluster already for other applications.  Docker containers are used within Kubernetes clusters to hold native code, for which Kubernetes creates a framework around network connections and scaling of resources up and down.  An introduction on using [Kubernetes with R is published at this blog post](https://code.markedmondson.me/r-on-kubernetes-serverless-shiny-r-apis-and-scheduled-scripts/).
- [`AzureContainers`](https://cran.r-project.org/package=AzureContainers): Umbrella package for working with containers in Microsoft's Azure cloud. Provides interfaces to three Azure services: Container Registry, Container Instances (for running individual containers), and Kubernetes Services (for orchestrated deployments)

While Plumber provides the low-level infrastructure for turning a statistical model into a predictive service, for production purposes it is usually necessary to take into account considerations such as scalability, reliability and ease of management.
A single container is limited in the volume of requests it can service.
Furthermore, if the container goes down, or is occupied with processing a computationally intensive task, the service becomes unavailable until it is restarted.

These issues can be mitigated by deploying not a single container, but a _cluster_ of containers, orchestrated as a single deployment. The deployment system of choice at the time of writing is [Kubernetes](https://kubernetes.io).
This allows you to deploy a single image to a pool of servers, which can then serve requests in parallel. Client-side requests are made to the cluster ingress endpoint, which then redirects them to individual servers.
The orchestrator can identify cluster nodes that have failed, and restart them automatically; it can also detect when there is no capacity to service more requests, and start additional servers and containers as needed.

While these load-balancing and autoscaling properties of a Kubernetes cluster are invaluable, they are still limited by the hardware resources that it has access to. A cluster cannot scale out if it has run out of servers.
This issue in turn can be addressed by deploying not to on-premise servers but to the cloud, where effectively unlimited computing resources are available (as long as one can pay for them).
From a management perspective, cloud deployments have the additional advantage that hardware maintenance becomes the responsibility of the cloud provider, eliminating another pain point.
Orchestrator services are available from all the major cloud service providers, including Amazon Web Services (AWS), Google Cloud Platform, and Microsoft Azure.

An example of an R package for working with containers in the cloud is AzureContainers, which interacts with three Azure services: Container Instances (for running a single container), Container Registry (a private docker registry service), and Kubernetes Service (for Kubernetes on Azure).
AzureContainers provides a lightweight yet powerful interface for working with Resource Manager, the framework for deploying and managing arbitrary resources in Azure.
On the client side, it provides simple shells to the Docker, Kubectl and Helm tools that are commonly used for managing containers and Kubernetes clusters.

Here is some simple code that pushes a Docker image to an Azure Container Registry, and then creates a deployment using Azure Kubernetes Service.
This is a slightly modified version of the code from the "Deploying a prediction service with Plumber" AzureContainers vignette.
The code makes use of a yaml file to define the deployment and predictive service, which is standard practice with Kubernetes; the yaml can be seen in the aforementioned vignette.

```r
library(AzureContainers)
# create a resource group for our deployments
deployresgrp <- AzureRMR::get_azure_login()$
    get_subscription("subscription_id")$
    create_resource_group("deployresgrp", location="australiaeast")

### create a container registry
deployreg_svc <- deployresgrp$create_acr("deployreg")

# build our image (a random forest on the Boston housing data, available in the MASS package)
call_docker("build -t bos-rf .")

# upload the image to Azure
deployreg <- deployreg_svc$get_docker_registry()
deployreg$push("bos-rf")

### create a Kubernetes cluster with 2 nodes
deployclus_svc <- deployresgrp$create_aks("deployclus", agent_pools=aks_pools("pool1", 2))

# grant the cluster pull access to the registry
gr <- AzureGraph::get_graph_login()
aks_app_id <- deployclus$properties$servicePrincipalProfile$clientID
reg$add_role_assignment(gr$get_app(aks_app_id), "Acrpull")

# get the cluster endpoint
deployclus <- deployclus_svc$get_cluster()

# create and start the service
deployclus$create("bos-rf.yaml")
```

Once the service has been created, we can check on its status and obtain predictions:

```r
deployclus$get("deployment bos-rf")
#> Kubernetes operation: get deployment bos-rf  --kubeconfig=".../kubeconfigxxxx"
#> NAME      DESIRED   CURRENT   UP-TO-DATE   AVAILABLE   AGE
#> bos-rf    1         1         1            1           5m

deployclus$get("service bos-rf-svc")
#> Kubernetes operation: get service bos-rf-svc  --kubeconfig=".../kubeconfigxxxx"
#> NAME         TYPE           CLUSTER-IP   EXTERNAL-IP     PORT(S)          AGE
#> bos-rf-svc   LoadBalancer   10.0.8.189   xxx.xxx.xxx.xxx 8000:32276/TCP   5m

response <- httr::POST("http://xxx.xxx.xxx.xxx:8000/score",
    body=list(df=MASS::Boston[1:10,]), encode="json")
httr::content(response, simplifyVector=TRUE)
#> [1] 25.9269 22.0636 34.1876 33.7737 34.8081 27.6394 21.8007 22.3577 16.7812 18.9785
```

Note that Plumber, by itself, provides no authentication or security features; anybody who knows the cluster's IP address can access the predictive service.
At minimum, a basic authentication layer should be added to any Kubernetes deployment.
Standard Kubernetes procedure involves creating an ingress controller, for example with the nginx or traefik reverse proxy software, and one can then add application-specific authentication layers on top of that (for example, to authenticate with a user's Azure Active Directory or LDAP credentials).
More details can be found from most Kubernetes learning resources, and are outside the scope of this paper.

@@

@@

**\pkg{outsider}** ([https://antonellilab.github.io/outsider/](https://antonellilab.github.io/outsider/)) tackles the problem of integrating external programs into an R workflow \citep{bennett_outsider_2020}.
Installation and usage of external programs can be difficult, convoluted and even impossible if the platform is incompatible. 
Therefore \pkg{outsider} uses the platform-independent Docker images to encapsulate processes in _outsider modules_.
Each outsider module has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
Using only R functions, an end-user can install a module with the \pkg{outsider} package and then call module code to integrate a tool into their own workflow seamlessly.
The \pkg{outsider} package and module manage the containers and handle the transmission of arguments and the transfer of files to and from a container.
These functionalities also allow a user to launch module code on a remote machine via SSH, expanding the potential computational scale.
Outsider modules can be hosted code-sharing services, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.
<!-- informative examples of how modules work, just as a reminder for @nuest
https://github.com/DomBennett/om..astral/blob/master/examples/astral.R
https://github.com/DomBennett/om..raxml#partitioned-dna-analysis
-->

@@

The ordering of authors following DN as first author is alphabetical. DN conceived the article idea, \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised all sections.
DB wrote the section on \pkg{outsider}.
DE wrote the introduction and the section about Containerization and Rocker.
RC \& EH contributed to the section on interfaces for Docker in R.
DC contributed content on Gigantum.
ME contributed to the section on processing and deployment to cloud services.
CF wrote paragraphs about \pkg{r-online}, \pkg{dockerfiler}, \pkg{r-ci} and \pkg{r-db}.
BM wrote the section on research compendia and made the project Binder-ready.
HN \& JN co-wrote the section on enterprise production environments (\pkg{googleComputeEngineR}, Google Cloud Run, Kubernetes).
KR wrote the section about \pkg{wholepunch}.
NR wrote paragraphs about shared work environments and GPUs.
LS \& NT wrote the section on Bioconductor.
NX wrote the section on \pkg{liftr}.
All authors approved the final version. This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
The \href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page}, \href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history}, and \href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues} provide a detailed view on the respective contributions.

>>>>

title: "The Rockerverse: Images, Packages, and Applications for Containerization with R"

@@

  - name: Gergely Daroczi
    # https://github.com/daroczig
      - \orcid{0000-0003-3149-8537}
    email: daroczig@rapporter.net

@@

  - name: Charlotte Van Petegem
    # https://github.com/charvp
      - Department WE02
      - Krijgslaan 281, S9
      - 9000 Gent, Belgium
      - \orcid{0000-0003-0779-4897}
    email: charlotte.vanpetegem@ugent.be

@@

The [**Jupyter Docker Stacks**](https://github.com/jupyter/docker-stacks/) project are a set of ready-to-run Docker images containing Jupyter applications and interactive computing tools \citep{project_jupyter_jupyter_2018}.
The `jupyter/r-notebook` image includes R and "popular packages", and naturally also the IRKernel ([https://irkernel.github.io/](https://irkernel.github.io/)), an R kernel for Jupyter so that Jupyter Notebooks can contain R code cells. 
R is also included in the catchall `jupyter/datascience-notebook` image\footnote{\href{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}}.
For example, these images allow users to quickly start a Jupyter Notebook server locally or build their own specialised images on top of stable toolsets.
R is installed using the Conda package manager, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See \texttt{jupyter/datascience-notebook}'s \texttt{Dockerfile} at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}.}.

@@

**Gigantum** ([http://gigantum.com/](http://gigantum.com/) is a platform for open and decentralized data science with a focus on using automation and user-friendly tools for easy sharing of reproducible computational workflows.
Gigantum builds on the _Gigantum Client_ (running either locally or on a remote server) for development and execution of data-focused _Projects_, which can be stored and shared via the _Gigantum Hub_ or via a zipfile export.
The Client is a user friendly interface to a backend using Docker containers to package, build, and run Gigantum projects. 
It is configured to use a default set of Docker base images ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)), and users are able to define and configure their own custom images.
The available images include two with R based on Ubuntu Linux have the [`c2d4u` CRAN PPA](https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/) pre-configured for installation of binary R packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
The R images vary in the included authoring environment, i.e., Jupyter in `r-tidyverse` or both Jupyter \& RStudio in `rstudio-server`.
The independent image stack can be traced back to the Gigantum environment and its features.
The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.
The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the \texttt{Dockerfile} of \texttt{python3-minimal}.} so that appropriate GPU drivers can be enabled automatically when supported.
Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resolution.
For this reason, project repositories store authoritative package information in a separate file per package, allowing Git to directly detect conflicts and changes.
A Dockerfile is generated from this description that inherits from the specified base image, and additional custom Docker instructions may be appended by users, though Gigantum's default base images do not currently include the `littler` tool, which is used by Rocker to install packages within `Dockerfile`s.
Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusion}).
Both approaches enable `apt` as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on plain `Dockerfile`s.
A minor reason in the inception of the images were also project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.

@@

**\pkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) is currently the most powerful Docker client in R.
It interfaces with the Docker daemon over the Docker HTTP API\footnote{} via a Unix socket on Linux or MacOS, over a named pipe on Windows, or over an HTTP/TCP connection.
The package is the only one not using system calls to the `docker` CLI for managing images and containers and easily exposes connections to remote Docker daemons, which has to be configured on the Docker level otherwise.
Using the API gives access to more information and is system independent and likely more reliable than parsing command line output.
\pkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but still similar to the Docker CLI.
The interface is similar to R6 objects, in that a `stevedore_object` has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/tree/master/inst/spec}{https://github.com/richfitz/stevedore/tree/master/inst/spec} and \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.

**\pkg{AzureContainers}** ([https://cran.r-project.org/package=AzureContainers](https://cran.r-project.org/package=AzureContainers)) is an interface to a number of container-related services in Microsoft's [Azure Cloud](https://azure.microsoft.com/), see Section&nbsp;\nameref{azurecontainers}.
While it is mainly intended for working with Azure, as a convenience feature it includes lightweight, cross-platform shells to Docker and Kubernetes (tools `kubectl` and `helm`).
These can be used to create and manage arbitrary Docker images and containers, as well as Kubernetes clusters on any platform or cloud service.

@@

R has been historically viewed as a tool for analysis and scientific research--not for creating software that corporations can rely on to continously run in real time.
However, thanks to advancements in R running as a web service, along with with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
One of the first implementations of having R respond to client reqests were TCP/IP and local socket servers, such as \CRANpkg{Rserve} and \CRANpkg{svSocket}, which allowed other local or remote processes to make binary requests to interact with R without the need to link to R at compile time and connected other programming languages with R.
R could also empower web services since at least 2007, when [rApache](http://www.rapache.net) was presented at the useR! conference \citep{horner_rapache_2013}.
<!-- website seems down, https://web.archive.org/web/20190627101203/http://rapache.net/manual.html -->
rApache runs an R interpreter inside the Apache web server, which makes it possible and very efficient to build scalable R-based APIs and web services.

Such rApache applications have been used at [rapporter.net](http://rapporter.net), transforming parametrized R Markdown-like documents into interactive web applications for reporting.
A similar design was implemented at [CARD.com](https://www.card.com/), where the optimization of the acquisition flow and the real-time analysis of debit card transactions both relied on an efficient combination of Rserve and rApache running in Docker containers:
the R session behind Rserve acted as a read-only, exteremely fast and scalabale, in-memory database for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
Similar dockerized R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time based on card transactions.
The related Docker images were made available at [https://github.com/cardcorp/card-rocker](https://github.com/cardcorp/card-rocker).
The images extend `rocker/r-base`, e.g., with an SSH client and a workaround for being able to mount SSH keys from the host, Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required by the AWS SDK.
The AWS SDK allowed to run R consumers reading from real-time data processing streams of [AWS Kinesis](https://aws.amazon.com/kinesis/).
The applications were deployed on Amazon Elastic Container Service ([ECS](https://aws.amazon.com/ecs/)).
The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to an LTS distribution to minimize the risk of breaking the automated builds of the Docker images and production jobs.

A more recent related work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package, which makes it even easier for R to be used to host web services through RESTful APIs.
When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.
For instance, if a data scientist at a company creates a machine learning model to tell which customers are likely to make a repeat purchase, if they serve the model through a plumber API then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
Plumber syntax is simple to read and understand: the API is created by commenting above function names in an R script.

@@

[RestRserve](http://restrserve.org) is an alternative R web API framework with the main focus on being high-performant and robust.
It also ships automated Docker builds at https://hub.docker.com/r/rexyai/restrserve/dockerfile.

@@

- `RSelenium`
- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `gce_vm_template()`)
  - To enable quick deployments of key R services such as RStudio and Shiny onto cloud virtual machines (VMs), this package utilises Dockerfiles to move the labour of setting up those services from the user to a premade Docker image.  For example, by specifying the template `template="rstudio"` in `gce_vm_template()/gce_vm()` an up to date RStudio Service image is launched.  Specifying `template="rstudio-gpu"` will launch an RStudio Server image with a GPU attached, etc.  
- [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
- [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]
- Kubernetes - A popular platform for managing Docker containers is [Kubernetes](https://kubernetes.io/), which is used for a wide variety of containerized applications.  It may be your organisation has a Kubernetes cluster already for other applications.  Docker containers are used within Kubernetes clusters to hold native code, for which Kubernetes creates a framework around network connections and scaling of resources up and down.  An introduction on using [Kubernetes with R is published at this blog post](https://code.markedmondson.me/r-on-kubernetes-serverless-shiny-r-apis-and-scheduled-scripts/).

\label{azurecontainers}
**\pkg{AzureContainers}** is an umbrealla package providing interfaces the three services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
While a package like \pkg{plumber} provides the infrastructure for turning an R workflow a service, for production purposes it is usually necessary to take into account scalability, reliability and ease of management.
A single container is limited in the volume of requests it can service, and if the container goes down or is occupied with processing a computationally intensive task, the service becomes unavailable.
These issues can be mitigated by deploying not a single container, but a _cluster_ of containers, orchestrated as a single deployment, e.g., with [Kubernetes](https://kubernetes.io).
A cluster features load-balancing, autoscaling of containers across numerous servers (in the cloud or on premise), and restarting failed ones.
\pkg{AzureContainers} provides a lightweight yet powerful interface for working with Resource Manager, the framework for deploying and managing arbitrary resources in Azure.
On the client side, it provides simple shells to the Docker, and to Kubernetes' `kubectl` and `helm` tools, that are needed for managing Kubernetes clusters.
The package simplifies complex infrastructure management to a number of R funcction calls, given an Azure account with sufficient credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.

@@

Another way this idea of a common environment and the partitioning that docker provides is used in teaching is in the **[Dodona](https://dodona.ugent.be)** project.
Dodona is a web platform developed at Ghent University and is used to teach students basic programming skills.
Dodona uses Docker containers to test submissions by students.
This means that both the code testing the students' submissions and the submission itself are executed in a predictable environment, avoiding compatibility issues between the wide variety of configurations used by students.
The containerization is also used to shield the Dodona servers from bad (or even malicious) code: memory, time and I/O limits are used to make sure students can't overload the system.
The web application communicates with containers by sending configuration information as a JSON document over standard input. 
Every Dodona docker image shares a `main.sh` file that passes through this information to the actual testing framework, while setting up some error handling. 
The testing process in the docker containers sends back the test results by writing a JSON document to its standard output channel.
In june 2019 R support was added to Dodona using an image derived from the Rocker `r-base` image that sets up the `runner` user and and `main.sh` file expected by Dodona\footnote{\href{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}}. 
It also installs the packages required for the testing framework and the exercises so that this doesn't have to happen every time a student's submission is evaluated.
The actual testing of R exercises is done using a custom framework loosely based on `testthat`.
During the development of the testing framework it was found that the testthat framework did not provide enough information to its reporter system to send back all the fields required by Dodona to render its feedback.
Right now, multiple statistics courses are developing exercises to automate the feedback for their lab classes.

@@

**\pkg{outsider}** ([https://antonellilab.github.io/outsider/](https://antonellilab.github.io/outsider/)) tackles the problem of integrating external programs into an R workflow \citep{bennett_outsider_2020}.
Installation and usage of external programs can be difficult, convoluted and even impossible if the platform is incompatible. 
Therefore \pkg{outsider} uses the platform-independent Docker images to encapsulate processes in _outsider modules_.
Each outsider module has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
Using only R functions, an end-user can install a module with the \pkg{outsider} package and then call module code to integrate a tool into their own R-based workflow seamlessly.
The \pkg{outsider} package and module manage the containers and handle the transmission of arguments and the transfer of files to and from a container.
These functionalities also allow a user to launch module code on a remote machine via SSH, expanding the potential computational scale.
Outsider modules can be hosted code-sharing services, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.
<!-- informative examples of how modules work, just as a reminder for @nuest
https://github.com/DomBennett/om..astral/blob/master/examples/astral.R
https://github.com/DomBennett/om..raxml#partitioned-dna-analysis
-->

@@

The ordering of authors following DN as first author is alphabetical. DN conceived the article idea, \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised all sections.
DB wrote the section on \pkg{outsider}.
GD contributed the CARD.com use case.
DE wrote the introduction and the section about Containerization and Rocker.
RC \& EH contributed to the section on interfaces for Docker in R.
DC contributed content on Gigantum.
ME contributed to the section on processing and deployment to cloud services.
CF wrote paragraphs about \pkg{r-online}, \pkg{dockerfiler}, \pkg{r-ci} and \pkg{r-db}.
BM wrote the section on research compendia and made the project Binder-ready.
HN \& JN co-wrote the section on enterprise production environments (\pkg{googleComputeEngineR}, Google Cloud Run, Kubernetes).
KR wrote the section about \pkg{wholepunch}.
NR wrote paragraphs about shared work environments and GPUs.
LS \& NT wrote the section on Bioconductor.
NX wrote the section on \pkg{liftr}.
All authors approved the final version. This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
The \href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page}, \href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history}, and \href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues} provide a detailed view on the respective contributions.
